---
title: "btt-discrete-time"
output: html_document
date: "2024-06-06"
---

## Load data and necessary libraries
## Dataset can be constructed from the heart_pipeline_final.Rmd file
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Molly:
setwd("C:/Users/mhwhi/Downloads")
load("heart_post_policy (4).RData")
df_cand_thor <- haven::read_sas('./pubsaf2403/cand_thor.sas7bdat')

# Grace:
# load("heart_post_policy.RData")
# curr_data_release = "//srtr-fs-p01/SAF/External/Pub_SAF/saf2403_Q1/pubsaf2403"
# df_cand_thor <- haven::read_sas(file.path(curr_data_release,'cand_thor.sas7bdat'))

library(renv)
library(dplyr)
library(tidyr)
library(lubridate)

# Subset to cohort
df2 <- subset(df, df$t_start == '0' & as.Date(df$unique_date) >= '2019-01-01' & 
                as.Date(df$unique_date) <= '2022-12-31')
postpolicy_id <- unique(df2$PX_ID)
rm(df2)
df <- df[which(df$PX_ID %in% postpolicy_id), ]

# Retain listing date, transplant date for waiting time calculations
df = df %>% mutate(can_listing_dt = if_else(grepl("CAN_LISTING_DT", unique_event), unique_date, NA),
                   transplant_date = if_else(grepl("REC_TX_DT", unique_event), unique_date, NA)) %>% fill(can_listing_dt, transplant_date, .direction = "downup")
```

## Censor candidates past December 31st, 2023
```{r}
df = df  %>% 
  filter(unique_date <= mdy("12-31-2023")) %>% 
  group_by(PX_ID) %>%
  mutate(
    t_stop = ifelse(row_number() == n() & t_stop - t_start >  mdy("12-31-2023") - as.Date(unique_date), t_start + mdy("12-31-2023") - as.Date(unique_date), t_stop)
  )
# we have recoded t_stop for candidates censored on 12/31/2023 but have not recoded outcome to be 0 if the death date is after 12/31/2023

```

## Identify candidates who were delisted
## Determine time to death / censor for these candidates
```{r}
df = df %>% mutate(
  delisting = ifelse(grepl("CAN_REM_DT", unique_event) & !grepl("REC_TX_DT", unique_event), 1, 0)
) %>% mutate(
  time_until_death_after_delisting = ifelse(delisting == 1 & outcome == 1, t_stop - t_start, NA),
  time_until_censor_after_delisting = ifelse(delisting == 1 & outcome == 0, t_stop - t_start, NA)) %>% group_by(PX_ID) %>% fill(time_until_death_after_delisting, time_until_censor_after_delisting, .direction = "up")

```


## Define short term mcs ever variables
```{r}

df <- df %>%
  group_by(PX_ID) %>%
  
  mutate(ECMO_ever = case_when(ECMO == '1' ~ '1', TRUE ~ NA_character_),
         temp_surg_ever = case_when(temp_surg_LVAD == '1' ~ '1', TRUE ~ NA_character_),
         BiVAD_no_discharge_ever = case_when(BiVAD_no_discharge == '1' ~ '1', TRUE ~ NA_character_)) %>%
  
  fill(ECMO_ever, .direction = 'down') %>%
  fill(temp_surg_ever, .direction = 'down') %>%
  fill(BiVAD_no_discharge_ever, .direction = 'down') %>%

  mutate(ECMO_ever = case_when(is.na(ECMO_ever) ~ '0', TRUE ~ ECMO_ever),
         temp_surg_ever = case_when(is.na(temp_surg_ever) ~ '0', TRUE ~ temp_surg_ever),
         BiVAD_no_discharge_ever = case_when(is.na(BiVAD_no_discharge_ever) ~ '0', TRUE ~ BiVAD_no_discharge_ever))


```

## Discrete time data construction
```{r}
# create tx dataset
# remove rows at time of transplant, removal
df_tx = df %>% filter(transplant == 1 | (CAN_REM_CD == 4 & grepl("CAN_REM_DT",unique_event)))
df = df %>% filter(cumsum(grepl("CAN_REM_DT", unique_event))<1)

# define period to group observations that occur within the same week
# take first observation within each week
# duplicate each observation based on how many days there are until the next period
# define interval_start, interval_stop as the 7 day periods

df_final = df %>% mutate(
  period = ceiling(t_stop/7)
) %>% group_by(PX_ID, period) %>% 
  # taking the first row per week
  filter(row_number() == 1) %>% 
  mutate(
    t_start = ceiling(t_start/7)*7,
    t_stop = ceiling(t_stop/7)*7, 
    duplicates = (t_stop - t_start)/7
  ) %>% ungroup() %>% uncount(duplicates) %>%
  group_by(PX_ID) %>%
  mutate(interval_start = seq(0, max(t_stop) - 7, by = 7),
                             interval_stop = interval_start + 7) %>%
  ungroup()

df_final

```






## Set transplant = 1 for PX_ID in transplant list, at their last observation on the waitlist
```{r}
tx_patients = unique(df_tx$PX_ID)

df_final = df_final %>% group_by(PX_ID) %>%
  mutate(
  transplant = ifelse(row_number() == n() & PX_ID %in% tx_patients, 1, 0)
) %>% select(-c(t_start, t_stop, period, unique_event, unique_date)) %>% 
  relocate(interval_start, interval_stop, .after=CAN_LISTING_CTR_CD)

df_final %>% filter(PX_ID == 1310477) %>% select(PX_ID, transplant)
```


# Body surface area calculation
```{r}
# select necessary cand thor variables
df_cand_thor <- df_cand_thor %>% select(PX_ID, CAN_GENDER, CAN_RACE, CAN_HGT_CM, CAN_WGT_KG)

# impute missing height and weights
df_cand_thor$CAN_HGT_CM[is.na(df_cand_thor$CAN_HGT_CM)] <-median(df_cand_thor$CAN_HGT_CM, na.rm=T)
df_cand_thor$CAN_WGT_KG[is.na(df_cand_thor$CAN_WGT_KG)] <-median(df_cand_thor$CAN_WGT_KG, na.rm=T)

# define body surface area, merge with discrete time data
df_cand_thor <- df_cand_thor %>%
  mutate(bsa = 0.007184 * (CAN_HGT_CM ^ 0.725) * (CAN_WGT_KG ^ 0.425)) 
df_cand_thor <- df_cand_thor[!duplicated(df_cand_thor), ]
df_final <- merge(df_final, df_cand_thor, by='PX_ID', all.x=T, all.y=F)


```



## Fill missing arterial lactate, LDH, Hemoglobin, BNP, sodium, albumin, bilirubin, creatinine
```{r}
# fill missing arterial lactate, LDH, Hemoglobin, BNP, sodium, albumin, bilirubin, creatinine
df_final <- df_final %>%
  mutate(
    arterial_lactate = ifelse(is.na(arterial_lactate), 1, arterial_lactate),
    LDH = ifelse(is.na(LDH), 220, LDH),
    HemoHemoglobin = case_when(
      is.na(HemoHemoglobin) & CAN_GENDER == 'M' ~ 14.9,
      is.na(HemoHemoglobin) & CAN_GENDER == 'F' ~ 13.3,
      TRUE ~ HemoHemoglobin
    ),
    BNP = case_when(
      is.na(BNP) & BNP_NT_Pro == '0' ~ median(BNP[BNP_NT_Pro == '0'], na.rm = TRUE),
      is.na(BNP) & BNP_NT_Pro == '1' ~ median(BNP[BNP_NT_Pro == '1'], na.rm = TRUE),
      TRUE ~ BNP
    ),
    sodium = ifelse(is.na(sodium), median(sodium, na.rm = TRUE), sodium),
    albumin = ifelse(is.na(albumin), median(albumin, na.rm = TRUE), albumin),
    bilirubin = ifelse(is.na(bilirubin), median(bilirubin, na.rm = TRUE), bilirubin),
    creatinine = ifelse(is.na(creatinine), median(creatinine, na.rm = TRUE), creatinine)
  )
  

```

## Fill missing cardiac_output, central_venous_pressure, PASP, PADP, PCWP, systolicBP, diastolicBP, resting_HR
```{r}
df_final <- df_final %>%
  ungroup() %>%
  mutate(
    cardiac_output = case_when(
      is.na(cardiac_output) & durable_LVAD == '1' ~ 2.5 * bsa,
      is.na(cardiac_output) & durable_LVAD != '1' ~ 2.2 * bsa,
      TRUE ~ cardiac_output
    ), 
    central_venous_pressure = case_when(
      is.na(central_venous_pressure) & durable_LVAD == '1' ~ 10,
      is.na(central_venous_pressure) & durable_LVAD != '1' ~ 12,
      TRUE ~ central_venous_pressure
    ),
    PASP = case_when(
      is.na(PASP) & durable_LVAD == '1' ~ 35,
      is.na(PASP) & durable_LVAD != '1' ~ 38,
      TRUE ~ PASP
    ),
    PADP = case_when(
      is.na(PADP) & durable_LVAD == '1' ~ 15,
      is.na(PADP) & durable_LVAD != '1' ~ 18,
      TRUE ~ PADP
    ),
    PCWP = case_when(
      is.na(PCWP) & durable_LVAD == '1' ~ 15,
      is.na(PCWP) & durable_LVAD != '1' ~ 18,
      TRUE ~ PCWP
    ),
    systolicBP = case_when(
      is.na(systolicBP) & durable_LVAD == '1' ~ 100,
      is.na(systolicBP) & durable_LVAD != '1' ~ 110,
      TRUE ~ systolicBP
    ),
    diastolicBP = ifelse(is.na(diastolicBP), 80, diastolicBP),
    resting_HR = ifelse(is.na(resting_HR), 8.5, resting_HR)
  )
```

## Calculate eGFR, define short mcs ever
```{r}
df_final <- df_final %>%
  mutate(eGFR = case_when(
    CAN_GENDER == 'F' & !is.na(creatinine) ~ 142 * (pmin((creatinine / 0.7), 1)^(-0.241)) *
      (pmax((creatinine / 0.7), 1)^(-1.2)) * 0.9938^(age_at_listing) * 1.012,
    
    CAN_GENDER == 'M' & !is.na(creatinine) ~ 142 * (pmin((creatinine / 0.9), 1)^(-0.302)) *
      (pmax((creatinine / 0.9), 1)^(-1.2)) * 0.9938^(age_at_listing)),
      short_MCS_ever = 
           case_when(ECMO_ever == '1' | temp_surg_ever == '1' | BiVAD_no_discharge_ever == '1' ~ 1,
             TRUE ~ 0))

df_final %>% filter(is.na(creatinine))
```
## Define hemodynamics, impute missing values
```{r}
# define hemodynamics, impute missing values
df_final <- df_final %>%
  mutate(
    cpo = (1/541) * cardiac_output * (((2/3)*systolicBP) + ((1/3)*diastolicBP)),
    api = (systolicBP - diastolicBP) / PCWP,
    papi = (PASP - PADP) / central_venous_pressure) %>%
  
  mutate(cpo = as.numeric(cpo),
         api = as.numeric(api),
         papi = as.numeric(papi)) %>%
  mutate(api = ifelse(PCWP == 0, 1, api),
         papi = case_when(
            central_venous_pressure == 0 ~ 1,
            papi < 1 & !is.na(papi) ~ 1,
            TRUE ~ papi
    ))
 


```

# Add time since waitlisting variable
```{r}
df_final = df_final %>% mutate(
  time_since_listing = interval_start
)
```


# Save discrete time data
```{r}
save("df_final", file =  "discrete_final_data.RData")
```



```{r}
df %>% filter(PX_ID == 1511673)
```








